---
title: "STT 301 Day 4 In-class work"
author: "Andy Finley"
date: "September 14, 2016"
output: html_document
---

First we will set code chunk options for the entire document. These can be overridden by the options in an individual code chunk, i.e., placed after the `r` in `{r }`. Typically we would use `echo = FALSE` in this chunk, since we wouldn't want it seen in the final document, but here we display the chunk.

```{r global_options}
knitr::opts_chunk$set(comment = NA, tidy = TRUE)
```

Today we will get practice with vectors in R. We'll read in three vectors, one that lists the teams who won the World Series, one that lists the teams who lost the World Series, and a third that gives the years the World Series was played. 

```{r}
con <- url("http://blue.for.msu.edu/STT301/data/WorldSeries.gz")
load(con)
close(con)
rm(con)
ls()
```

List the first 50 World Series winning teams.

```{r}
#List the winning teams here
```

List the first 50 World Series losing teams.
```{r}
#List the losing teams here
```

List the first 50 years that the World Series was played.

```{r}
#List the years here
```

The `table` function builds a contingency table. For example,

```{r}
animals <- c("dog", "cat", "dog", "cat", "horse", "dog")
table(animals)
```

Create a table which gives the number of times each team has won the World Series.

```{r}
#Show the table here
```

By default the table is ordered alphabetically by the name of the team. It might be more useful to order by the number of World Series won. Apply the `sort` function to the table to achieve this, and save the resulting table as `winners_table`.

```{r}
#Show the sorted table here
```

Which two teams have won the most World Series?

Use the `[]` subsetting capabilities of R applied to `winners_table` to create a table called `top_winners` which contains the five teams which won the World Series most often. (There are several teams which are tied with five wins, and two of these will show up in the fourth and fifth positions. Don't worry about the fact that we should include the other teams too.)

```{r}
###Create the top_winners table here
```

Draw a bar graph of the `top_winners` table using the `barplot` function.

```{r}
#Create the bar graph here
```

It's likely that the labels for some of the bars did not fit on the bar graph you drew. The argument `cex.names` allows you to increase (with values larger than 1) or decrease (with values less than 1) the size of the font. Experiment with `cex.names` to get the labels all to fit on the graph.

```{r}
#Create the modified bar graph here
```


Now create a table of the losing teams, and sort in decreasing order, by giving the `decreasing = TRUE` argument to `sort`.


```{r}
#Show the sorted table here
```

Which two teams have lost the World Series most often?

Create a bar graph of the five teams which have lost most often, again making sure the labels are all printed.

```{r}
#Create the bar graph here.
```

Use R to find out how many years has the World Series been played.
```{r}
#Compute the answer here
```

Use R to list the most recent World Series winner. Do this in a way that does NOT require you to know how many World Series there have been.

```{r}
#List the most recent winner here.
```

Use R to list the past eight World Series LOSING teams. Again, you shouldn't need to know how many World Series have been played.

Find the type of each of the three vectors.
```{r}
#List the types here.
```

The vectors listing the winning and losing teams would be more informative if the years of the wins and losses were included. We can give names to each vector element using the `names` function. For example:
```{r}
animals
names(animals) <- c("Asta", "Felix", "Chopper", "Garfield", "Mr. Ed", "Lassie")
animals
```

Use `names` to add the years of the World Series as names for both the `winners` and `losers` vectors, and list the first ten elements of each vector with the names included.
```{r}
#Create the vectors with names and list the first ten elements of each here.
```

Create a vector called `sorted_winners` which gives the winning teams sorted alphabetically. List the first thirty teams in this vector, and find (by inspection) the years when the Cleveland Indians won the World Series.

```{r}
#Create the sorted vector here and list the first 30 elements of the vector
```

###More advanced subsetting

So far we have directly specified the elements of vectors that we want to extract, for example `winners[1]` or `losers[length(losers)]` or `winners[c(1,2,3,5)]`. More typically we want to extract elements that meet a condition, such as all the winners before 1950, or all the losers during years when the New York Yankees won the World Series. For this we use subsetting with logical vectors. This is formally discussed in Section 4.7 of the notes, but we will informally use it now.

First some examples with very simple vectors, and then you'll use these ideas for the World Series vectors.

```{r}
set.seed(123) #Set the seed for reproducible random number generation
x <- rbinom(10, 20, 0.9) #Generate 10 binomial(20, 0.9) values
y <- rbinom(10, 20, 0.8) #Generate 10 binomial(20, 0.8) values
x
y
x[x > 16] #Values of x which are larger than 16
x[x == 18] #Values of x which are equal to 18
length(x[x == 18]) #How many values of x are equal to 18?
x[y < 15] #Values of x for which the corresponding values of y are less than 15
x[x < y] #Values of x which are less than the correponding values of y
x[x == 18] <- 0 #Replace 18 by 0 in x
x
```

Now we use logical subsetting to learn about the World Series data. 
First, use logical subsetting to count the number of World Series won by the Detroit Tigers, and also to count the number of World Series lost by the Detroit Tigers.

```{r}
#Count Tigers winners and losers here
```

Next use logical subsetting to determine the years when the Baltimore Orioles won the World Series.

```{r}
#Years that Baltimore won here
```

Next use logical subsetting to determine the losing teams during the years when the St. Louis Cardianals won the World Series.

```{r}
#Losers to the Cardinals here
```

We can combine conditions with the operators `&` and `|` which represent "and" and "or". Also `!` negates, so for example `!=` represents not equal to. Also `x %in% y` is useful:

```{r}
letters
letters %in% c("a", "m", "q", "s")
```
Use these ideas to 

a. Extract the years when either the Boston Red Sox or the Chicago White Sox won the World Series.
b. Extract the years when the New York Yankees won the World Series and the New York Giants lost the World Series.
c. Extract the winners during the years 1903, 1923, 1943, 1963, 1983.
d. Count the number of years during which a team from Boston won the World Series. Be careful: There are three teams from Boston in the data set.

```{r}
#Answers to a-d here.
```

###Another data set
Data on the number, in thousands, of international airline passengers from 1949-1960 are built in to R. These are in the form of a time series, and we'll first convert these to a vector since that's the type of data structure we currently know how to work with.

```{r}
passengers <- as.vector(AirPassengers)
passengers
```

Investigate these data using R functions such as `mean`, `max`, `min`, `summary`, `sd`, `length`, etc. 
```{r}
#Show your preliminary explorations here
```

Draw a histogram of the number of passengers using the `hist` function.

```{r}
#Draw the histogram here
```

What do you notice?

Draw a scatter plot of the number of passengers, in order, using the `plot` function. The main arguments to `plot` are `x` and `y`, vectors of x and y values. First draw the scatter plot with no further arguments, and then add the argument `type = "l"` which asks R to connect the points, which is probably more informative for data such as these.
```{r}
#Draw the two scatter plots here
```

What do you  notice from the scatter plots?